---
title: "HW1"
author: "Zakari L. Billo"
date: "2026-02-25"
output: pdf_document
---

```{r setup, include=FALSE}
library(caret)
library(tidymodels)
library(tidyverse)
library(glmnet)

knitr::opts_chunk$set(
  fig.width = 8,
  fig.height = 7,
  fig.align = "center",
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

(a) Fit a lasso model on the training data. Report the selected tuning parameter and
the test error. When the 1SE rule is applied, how many predictors are included in
the model?

(b) Fit an elastic net model on the training data. Report the selected tuning parameters
and the test error. Is it possible to apply the 1SE rule to select the tuning parameters
for elastic net? If the 1SE rule is applicable, implement it to select the tuning
parameters. If not, explain why.

(c) Fit a partial least squares model on the training data and report the test error. How
many components are included in your model?

(d) Choose the best model for predicting the response and explain your choice.

(e) If R package “caret” was used for the lasso in (a), retrain this model using R package
“glmnet”, and vice versa. Compare the selected tuning parameters between the two
software approaches. Should there be discrepancies in the chosen parameters, discuss
potential reasons for these differences.

```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
